---
title: "Data Visualization Final Project"
format: revealjs
---

```{r}
# Put data prep work above
#| echo: false
#| output: false
#| warning: false
#| message: false
library(tidytuesdayR)
tuesdata <- tidytuesdayR::tt_load('2024-03-26')
team_results <- tuesdata$'team-results'
public_picks <- tuesdata$'public-picks'
library(readr)
cbb24 <- read_csv("cbb24.csv")
library(tidyverse)
library(shiny)
library(kableExtra)
# let's add a win percentage variable to cbb24
cbb24 <- cbb24 |> mutate(win_perc = W / G * 100)

# let's get all of the possible statistics choices we can look at from cbb24 (as a vector)
rs_stat_choices <- names(cbb24)[c(4:21, 23:24)]

pr_stats_choices <- names(team_results)[c(2:10)]

pr_rounds <- names(team_results)[c(11:16)]

all_teams <- cbb24 |> mutate(TEAM = as.factor(TEAM)) |>
  pull(TEAM) |> levels()

tourney_teams <- public_picks |> mutate(TEAM = as.factor(TEAM)) |>
  pull(TEAM) |> levels()

public_picks_rounds <- names(public_picks)[c(4:9)]
```

```{r}
#| echo: false
#| output: false
#| warning: false
#| eval: false
# For side by side plots
library(gridExtra)
grid.arrange(plot1, plot2, nrow = 1)
```


## Data Overview 1

  * The first data set we will be using is the `cbb24` data set. This data set contains various statistics from the 2023-2024 NCAA basketball regular season.
  
```{r}
cbb24_top_seeds <- cbb24 |> 
  filter(TEAM %in% c("Houston", "Connecticut", "Purdue", "North Carolina"))

cbb24_top_seeds[1:13] |>
  kable() |>
  kable_styling(full_width = FALSE,
                font_size = 23)

cbb24_top_seeds[14:24] |>
  kable() |>
  kable_styling(full_width = FALSE,
                font_size = 23)
```

## Data Overview 2

  * The second data set we will be using is the `public_picks` data set. This data set contains predictions made by the public as to how each team will do in the 2023-2024 NCAA tournament.
  
```{r}
public_picks |> filter(TEAM %in% c("Houston", "Connecticut", "Purdue", "North Carolina")) |>
  kable() |>
  kable_styling(full_width = FALSE,
                font_size = 25)
```

## Visualization 1

```{r}
## Visualization 1: cbb24 filtered for specific teams and specific stats
cbb24_top_teams <- cbb24 |> 
  filter(TEAM %in% c("Houston", "Connecticut", "Purdue", "North Carolina")) |>
  select(TEAM, win_perc, '2P_O', '2P_D', '3P_O', '3P_D')

cbb24_top_teams_long <- pivot_longer(cbb24_top_teams, 
                                     cols = -TEAM, 
                                     names_to = "Variable", 
                                     values_to = "Value")

ggplot(cbb24_top_teams_long, aes(x = Variable, y = Value, fill = TEAM)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Variable", y = "Value",
       title = "Performance Comparison of Top College Basketball Teams",
       subtitle = "Data from 2023-2024 NCAA Basketball Regular Season") +
  theme_minimal(base_size = 20) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_viridis_d()
```

## Visualization 2

```{r}
## Visualization 2: public picks filtered for specific teams and specific stats

## First, we need to convert the percentages to numeric
public_picks[, c("R64", "R32", "S16", "E8", "F4", "FINALS")] <- 
  lapply(public_picks[, c("R64", "R32", "S16", "E8", "F4", "FINALS")], 
         function(x) as.numeric(sub("%", "", x)))

rounds_order <- c("R64", "R32", "S16", "E8", "F4", "FINALS")

public_picks_top_teams <- public_picks |> 
  filter(TEAM %in% c("Houston", "Connecticut", "Purdue", "North Carolina")) |>
  select(TEAM, R64, R32, S16, E8, F4, FINALS)

public_picks_top_teams <- public_picks |> 
  filter(TEAM %in% c("Houston", "Connecticut", "Purdue", "North Carolina")) |>
  select(TEAM, S16, E8, F4, FINALS)

public_picks_top_teams_long <- pivot_longer(public_picks_top_teams, 
                                     cols = -TEAM, 
                                     names_to = "Round", 
                                     values_to = "Percentage")

# Recode the "Round" column as ordered factor
public_picks_top_teams_long$Round <- factor(public_picks_top_teams_long$Round, levels = rounds_order)

ggplot(public_picks_top_teams_long, aes(x = Round, y = Percentage, fill = TEAM)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Round", y = "Percentage",
       title = "Public Picks for 2024 NCAA Tournament Teams") +
  theme_minimal(base_size = 20) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_viridis_d()
```

## Visualization 3

  * Currently working on a visualization that will show stats for all team results from previous NCAA tournaments
